---
title: "Module 9 Notes"
author: "Alexandra Boville"
date: "2024-01-31"
output: html_document
---

```{r}
library(sjmisc)
library(tidyverse)
```
5 number summary (John Tukey 1977)
- sample minimum
- lowest for first quartile
- median
- upper quartile or third quartile
- sample maximum
- mean/arithmetic average

```{r}
f <- "~/Desktop/BIO384K - Applied Data Analysis/Country-Data-2016.csv"
d <- read_csv(f, col_names = TRUE, show_col_types = FALSE)
names(d)
```

```{r}
attach(d)
mean(life_expect, na.rm = TRUE)
```
```{r}
detach(d)
#mean(life_expect, na.rm=TRUE) # this throws an error!
mean(d$life_expect, na.rm = TRUE)
```
NOTE: The with() function accomplishes much the same thing as attach() but is self-contained and cleaner, especially for use in functions. If you use with(), however, all code to be run should be included as an argument of the function.

```{r}
with(d, mean(life_expect, na.rm = TRUE))
```

##9.5 Exploring Single Variables
Loading data
```{r}
f <- "~/Desktop/BIO384K - Applied Data Analysis/Country-Data-2016.csv"
d <- read_csv(f, col_names = TRUE, show_col_types = FALSE)
head (d)
```
```{r}
names(d)
```

Summarizing number of variable observation and number of countries
```{r}
nrow(d)
ncol(d)
length(d$country)
```


```{r}
summary(d)
```
checking for duplicate country rows.... there are none
```{r}
c <- length(unique(d$country))
c
```

Mean area and population size for all countries
```{r}
median(d$area, na.rm = TRUE)
median(d$population, na.rm = TRUE)
```

Creating new density variable which is population/area. 
```{r}
density <- (d$population/d$area)
d_density <- d %>% 
  mutate(density)
#other ways to create density column
#d$density <- d$population/d$area
#d <-cbind(d, density) using density assigned to vector above
d_density

```

What are the 10 most dense countries?
```{r}
d <- d_density[order(d_density$density, decreasing = TRUE), ]
#alternative way to sort 
#d <- d |> arrange(desc(density))
#head(d,10)
#special pipe operator %<>% so it would automatically assign to d (re-assignnment)
#d %<>% arrange(desc(density))
d[1:10, ]
```
The 10 least dense?
```{r}
d <- d_density[order(d_density$density, decreasing = FALSE), ]
d[1:10, ]
```

